Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-04-01 pubs/tfp16/paper.tex
random ramblings

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline.
\end{itemize}

@@

\subsection{STG to C}
The back-end is written in Haskell and generates proper tail-calling C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which, of course, may be different than those for non-strict
semantics.  Sharing is preserved.

\subsubsection{Proper tail calls.}
While performance is not a high priority, we attain proper tail calling
instead of using the trampoline used in earlier versions of
GHC~\cite{PJ-stockhardware}.  Because we maintain our own data and control
(continuation) stack all generated C functions take no explicit arguments and
have return type \emph{void}, and with a single exception in the non-strict
case are invoked in tail call positions, at non-zero optimization levels both
gcc and Clang/LLVM generate jumps instead of calls for C calls.  When using
gcc we also use a fixed set of machine registers for passing/returning values.

The exceptions to tail calling are for convenience.  Similar to what is
described in \emph{EA} we have a generic variadic \emph{stgApply} function
that is used for applications of non-known functions, and known functions with
a deficit or excess number of the arguments.  The \emph{stgApply} function takes
as arguments the first element of the application (\emph{FUN}, \emph{PAP},
\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
evaluates it with a call (rather than jump) to yield a \emph{FUN} or
\emph{PAP}.  This allows us to implement a single \emph{stgApply} function as
a C function taking a contiguous block of machine words (unboxed values
or pointers) on the explicit stack.  The implementation of \emph{stgApply}
as described in \emph{EA} makes the same concession regarding
call-with-return, but the scheme described there uses an auto-generated family
of \emph{stgApply} functions parameterized by number and type (unboxed or
pointer) of arguments.  We used this scheme initially as well, but because the
parameter block contains a bitmap describing which arguments are boxed/unboxed
for the benefit of the garbage collector, this elaboration is unnecessary.
For \emph{direct calls}, in which a known function is applied to a number
of arguments equal to its arity, \emph{stgApply} need not be invoked:
the \emph{FUN} is tail-called directly.

Call-with-return becomes even more convenient when considering the various
forms of strictness: when evaluating function or constructor arguments it is
simpler, and arguably more efficient in terms of time (but not space in
general beca) to ``evaluate in place'' than implement the continuation style
to the logical limit.  We are aware, however, that we may need to eliminate
this use of the native C stack so that we can properly unwind the stack in a
more sophisticated parallel runtime when pruning multiple evaluations of the
same thunk, and we have determined that this is a relatively minor exercise:
one possibility is that \emph{stgApply} takes an additional argument
indicating which of its subsequent arguments have already been forced, and at
each invocation if unforced arguments remain, evaluates the next (possibly)
unevaluated argument and tail calls itself, re-using the stack frame with just
this new argument updated.

\emph{It seems that we can dispose with call-with-return of STG functions,
  including stgApply (as opposed to runtime functions) without too much
  effort.  It also seems like this is needed if we want to be able to unwind
  the stack as described in the ``grey hole'' paper.  I'll try to do this this
  week (3/31/16-4/1/16) to verify the feasibility before changing the
  preceding description (which isn't quite right)}

>>>>

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}

The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we discuss at some length.

@@

\subsection{STG to C}
The back-end is written in Haskell and generates C code;
the runtime system is written in C/C++.  The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which, of course, may be different than those for non-strict
semantics.  Sharing is preserved.

\subsubsection{Proper tail calls.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009}.  To achieve efficient execution they
identified the need for the ability to examine and unwind the stack.  This
requires that use of the C stack be highly constrained if we are to 
avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
non-portable code to directly manipulate the C stack, complicating an
already non-trivial implementation effort.  We therefore must limit the
use of the C stack to calls to the runtime, in other words, when executing
code generated from the user program (and not in a runtime call), there
must be no growing of (pushing return addresses or data onto) the C stack.

First we establish that this is possible.  By maintaining our own data
and control (continuation) stack all generated C functions can take no
explicit arguments and have return type \emph{void}. If we arrange that
generated C code has only calls (other than runtime calls, which do not call
back to user-program code) in tail call positions, that is, immediately before
a return in the control flow graph, then at non-zero optimization levels both
gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When using
gcc we can also use a fixed set of machine registers for passing/returning
values, much like GHC.

\subsubsection{A taste of the continuation style.}
Now to implementation, which we discuss for a simple case, then the most complex
case.  To clearly distinguish between a notional C call (a syntactic concept)
and a machine-level call (pushing a return address onto the native C stack) we
will refer to the latter as \emph{call-with-return}, and to a machine-level
tail call as \emph{jump}.

First a detail.  Our implementation of the STG abstract machine has an
abstract register called the \emph{current value register} or \emph{CVR}.
Roughly speaking, it can hold the pointer to what is about to be evaluated or
what was just evaluated.  Now consider the evaluation of \texttt{case e of
  <alts>} if call-with-return is disallowed.  If the compiler cannot determine
that evaluating \texttt{e} does not jump somewhere then there must be some
mechanism to get to the \texttt{alts}.  Our solution is to have a separate C
function for the \texttt{alts}, and before executing the code to evaluate
\texttt{e}, to push a continuation---the address of this function and the free
variables of the \texttt{alts}---onto the stack.  Whatever the code for
\texttt{e} might do, ultimately (in the absence of error or non-termination)
the value of \texttt{e} is in the CVR, and some code somewhere reaches its end
and ``returns''---jumps to the address
in the continuation on the top of the stack, thus entering the code for
\texttt{alts}.  Thus continuations direct where to go to, not how to return
to caller as in a stack for an imperative language.

\subsubsection{Is this too much for this paper?}
Similar to what is described in \emph{EA}, in an earlier
implementation we have a generic variadic \emph{stgApply} C function that is
used for applications of non-known functions, and known functions with a
deficit or excess number of the arguments.  The \emph{stgApply} function takes
as arguments the first element of the application (\emph{FUN}, \emph{PAP},
\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
evaluates it with a call-with-return to yield a \emph{FUN} or \emph{PAP}.
This allows us to implement a single \emph{stgApply} function as a C function
taking a contiguous block of machine words (unboxed values or pointers) on the
explicit stack.  The implementation of \emph{stgApply} as described in
\emph{EA} performs the same call-with-return, but the scheme described there
uses an auto-generated family of \emph{stgApply} functions parameterized by
number and type (unboxed or pointer) of arguments.  (We used this scheme
initially as well, but because the stack is now fully explicit and the stack
frame is self-describing (e.g., contains a bitmap describing which arguments
are boxed/unboxed for the benefit of the garbage collector), this elaboration
is unnecessary.) Call-with-return was also used in the case that a \emph{FUN}
or \emph{PAP} is provided more than its arity number of arguments, in which
case \emph{FUN} or \emph{PAP} is called-with-return with its arity number
of objects, and the returned object and remaining argument passed (via jump)
to \emph{stgApply}.  Finally, for strict evaluation orders it is straightforward
to evaluate function or constructor arguments one at a time with call-with-return.

At that point of development these calls-with-return were clearly a hack
in an otherwise continuation-passing (via the explicit stack) control flow.
Because call-with-return could be of unbounded call depth this would be
problematic for stack unwinding.

The solution, slightly simplified, entailed adding a new parameter to \emph{stgApply}
and make the first parameter, the object to be applied, implicit.  The implicit
parameter is in the CVR---a pointer to the object to be applied when jumping to \emph{stgApply}.
The new argument simply indicates which of the boxed arguments

For \emph{direct calls}, in which a known function is applied
to a number of arguments equal to its arity, \emph{stgApply} need not be
invoked: the \emph{FUN} is tail-called directly.

In addition to being in a tail call position, the gcc constraints for
generating a tail call (jump) are (approximately) that the total size of the
caller and callee arguments be equal, and the return type sizes be equal.  For
Clang the requirement is slightly more strict: caller and callee type
signatures must be the same.  These are a consequence of the C calling
convention that the caller, not callee, cleans up the stack (removes the
callee's stack frame).  However, because we maintain our own stack we can
define our own calling convention(s), so tail calls can be made to functions
of notionally differing type.  In particular, top-of-stack frames can be
adjusted in size and content without undue copying.
